Simulation of Congenital Heart Defect Corrective Surgeries Using Thin Shell Elements

### 2012-03-02 Doc/Miccai/shell-model.tex
Improved shell-model section.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@386 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
Improved simulation chapter. Added ref to fig 1.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@387 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Options on the choice of model are limited. Mass-spring models, often used for cloth simulation \cite{Volino2005b}, allow for fast computation but do not posses the required level of physical accuracy and therefore are inappropriate for predictive simulations. On the contrary, simulation models based on the formulation of finite elements, either 3-dimensional elements or thin shell elements, are commonly used when high physical accuracy is required but are well known to be computationally expensive. The hybrid blood vessel model of \cite{Li2009} strikes a balance but is too constrained to be used as a truly universal approach for all kinds of surgical procedures used in CHD corrective surgery. We propose a model based on thin shell elements that still allows for fast computation with a reasonable number of elements.

\begin{figure}[tbh]
  \centering
  \includegraphics[width=0.4\columnwidth]{img/shell.pdf}
  \caption{Shell element combines bending with in-plane deformations.}
  \label{fig-shell}
\end{figure}

Formulation of thin shell elements is based on the machanics of continuum. For a thorough description we refer the reader to a standard textbook about finite element methods (FEM), i.e. \cite{Reddy1993}. Thin shell elements are commonly used in physics and mechanics to model deformation of thin structures because in that case they are numerically more stable than volumetric FEMs. A shell element is a 2-dimensional element whose thickness is a parameter of the simulation. They combine forces from in-plane deformations, i.e. stretching or shearing, with bending forces. Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of the current shape $\myvec{x}$ to the rest shape $\myvec{x_0}$. To accelerate computation, we use a corotational formulation that is based on the definition of local frames for each element.

>>>>

Options on the choice of model are limited. Mass-spring models, often used for cloth simulation \cite{Volino2005b}, allow for fast computation but do not posses the required level of physical accuracy and therefore are inappropriate for predictive simulations. On the contrary, simulation models based on the formulation of finite elements, either 3-dimensional elements or thin shell elements, are commonly used when high physical accuracy is required but are well known to be computationally expensive. The hybrid blood vessel model of \cite{Li2009} strikes a balance but is too constrained to be used as a truly universal approach for all kinds of surgical procedures used in CHD corrective surgery. We propose to use a model based on thin shell elements. Such a model still allows for fast computation when using a reasonable number of elements during simulation.

\begin{figure}[tbh]
  \centering
  \includegraphics[width=0.4\columnwidth]{img/shell.pdf}
  \caption{TODO: Thin shell elements model bended plates.}
  \label{fig-shell}
\end{figure}

Formulation of thin shell elements is based on the machanics of continuum. For a thorough description we refer the reader to a standard textbook about finite element methods (FEM), i.e. \cite{Reddy1993}. Thin shell elements are commonly used in physics and mechanics to model deformation of thin structures because in that case they are numerically more stable than volumetric FEMs. A shell element is a 2-dimensional element whose thickness is a parameter of the simulation. They combine forces from in-plane deformations, i.e. stretching or shearing, with bending forces (see figure \ref{fig-shell}). Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of the current shape $\myvec{x}$ to the rest shape $\myvec{x_0}$. To accelerate computation, we use a corotational formulation that is based on the definition of local frames for each element.

### 2012-03-02 Doc/Miccai/simulation.tex
Improved simulation chapter. Added ref to fig 1.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@387 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

We present a new approach that overcomes these difficulties by performing topological changes on the mesh to connect shell elements along joining edges (see figure \ref{fig-}. Since there is a certain distance between two joining edges in general, a topological modification alters the shape and size of affected shell elements. Note that in this simulation state there is temporarily no correlation to reality. We then start a controlled relaxation process during which the shell elements progressivly converge to their energetical equilibrium. To keep the simulation stable we avoid a strong increase in internal forces by linearly interpolate the rest shape of the connected mesh to the initial rest shape before joining. The effect is similar to contracting springs but there is no such problem like surplus energy after the joining process is done.

>>>>

\begin{figure}[tbh]
\begin{center}
\includegraphics[width=\columnwidth]{img/new-joining.png}
\end{center}
\caption{Joining of thin shell element meshes. Left: Split meshes. Center: Shell elements along the joining edges are connected. Right: Final result after the simulation.}
\label{fig-JoiningVessels}
\end{figure}

We present a new approach that overcomes these difficulties by performing topological changes on the mesh to connect shell elements along joining edges (see figure \ref{fig-JoiningVessels}). Since there is a certain distance between two joining edges in general, a topological modification alters the shape and size of affected shell elements. Note that in this simulation state there is temporarily no correlation to reality. We then start a controlled relaxation process during which the shell elements progressivly converge to their energetical equilibrium. To keep the simulation stable we avoid a strong increase in internal forces by linearly interpolate the rest shape of the connected mesh to the initial rest shape before joining. The effect is similar to contracting springs but there is no such problem like surplus energy after the joining process is done.
